---
title: "stuff-plus"
author: "Bryan Mui, BSA Baseball Consulting"
date: "2025-03-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Library packages

```{r packages, include=FALSE}
library(tidyverse)
#library(xgboost)
```

# Read the dataset

```{r}
data_original <- read_csv("./data/UCLA2023-2024.csv")
# The data set that we will be mutating
data <- data_original
head(data, 25)
```


# Outputting the columns

```{r}
colnames(data)
```

Predictor Ideas:

Velocity:

    * RelSpeed (Release Speed)
    * ZoneSpeed (Speed at the plate)
    * EffectiveVelo (Velocity adjusted for approach angle)

Movement:

    * VertBreak (Vertical movement due to spin)
    * InducedVertBreak (More refined vertical movement measurement)
    * HorzBreak (Horizontal movement due to spin)
    * pfxx (Horizontal movement component)
    * pfxz (Vertical movement component)

Spin:

    * SpinRate (Total revolutions per minute)
    * SpinAxis (2D spin direction)
    * SpinAxis3dTransverseAngle (3D spin components)
    * SpinAxis3dLongitudinalAngle
    * SpinAxis3dActiveSpinRate
    * SpinAxis3dSpinEfficiency

Release & Extension:

    * RelHeight (Height of release)
    * RelSide (Side angle of release)
    * Extension (How far forward the pitcher releases the ball)

Pitch Type & Classification:

    * TaggedPitchType (Human-classified pitch type)
    * AutoPitchType (Algorithm-classified pitch type)

Location & Trajectory (Optional, but can improve Stuff+ models):

    * PlateLocHeight (Height of the pitch as it crosses the plate)
    * PlateLocSide (Side location at home plate)
    * VertApprAngle (Vertical approach angle)
    * HorzApprAngle (Horizontal approach angle)
    
    
**For now, focusing on these variables:**

    * Pitch Velocity
    * Vertical Break
    * Horizontal Break 
    * Arm Angle 
    * Release Extension
    
    
# Stuff+ 

## Part 1: Exploring Pitch Types and Sectioning Data Based off Pitches

```{r}
data %>% 
  group_by(TaggedPitchType) %>%
  summarize(Count = n())

data %>%
  mutate(Pitcher = as.factor(Pitcher))
```
We can see that we have ample data to produce a model for 1) Fastball, 2) Curve Ball, 3) Change Up, 4) Slider, 5) Sinker. The rest of the pitches have limited observations

    
## Part 2: Target Variable

Going to create a predicted hit probability predictor on a 0–1 scale, defined like:

* 0 = Home run
* 0.2 = Dangerous Contact
  - High exit velocity (e.g., > 95 mph)
  - Ideal launch angle (e.g., 10–30°)
* 0.5 = Weak contact/foul
  - Low exit velocity or bad angle (pop-ups, grounders, fouls)
* 0.8 = Called Strike
* 1 = Whiff/Swinging Strike


```{r}
# Calculate Hit or CSW Score
data <- data %>%
  mutate(
    hitCSW = case_when(
      PlayResult == "HomeRun" ~ 0,
      ExitSpeed > 95 & Angle >= 10 & Angle <= 30 ~ 0.1, # dangerous hit
      PitchCall %in% c("InPlay") ~ 0.3,
      PitchCall == "BallCalled" | PitchCall == "BallIntentional" | PitchCall == "BallinDirt" | PitchCall == "HitByPitch" | PitchCall == "Undefined"    ~ 0.6,
      PitchCall == "FoulBall" | PitchCall == "FoulBallFieldable" | PitchCall == "FoulBallNotFieldable"   ~ 0.6,
      PitchCall %in% c("StrikeCalled") ~ 0.8,
      PitchCall %in% c("StrikeSwinging") ~ 1, 
    )
  ) %>%
  relocate(hitCSW, .after = PitchCall)

# Predict Hit or CSW for per pitch(Using Regression)

# data %>%
#   group_by(PitchCall) %>%
#   summarize(Count = n())
# 
# data %>%
#   group_by(PlayResult) %>%
#   summarize(Count = n())
# 
# data %>%
#   group_by(hitCSW) %>%
#   summarize(Count = n())

```

    
### Normalizations

We need to apply the following normalizations:

* Count
* Approach Angle
* Spin Direction
    
```{r}
# Normalize Based on Count
count_model <- lm(hitCSW ~ Balls + Strikes, data = data)
summary(count_model)

data <- data %>%
  mutate(hitCSW_resid = resid(count_model)) %>%
  relocate(hitCSW_resid, .after = hitCSW)

data %>%
  select(Balls, Strikes, hitCSW_resid) %>%
  group_by(Balls, Strikes) %>%
  summarize(mean_hitCSW = mean(hitCSW_resid))
```



    
## Part 3: Calculating Stuff+

```{r}
# Select the variables we need, refer the beginning to see which variables are being selected
vars <- c(
    "Pitcher",
    "PitcherId",
    "TaggedPitchType",
    "RelSpeed",
    "ZoneSpeed",
    "EffectiveVelo",
    "VertBreak",
    "InducedVertBreak",
    "HorzBreak",
    "SpinRate",
    "SpinAxis",
    "RelHeight",
    "RelSide",
    "Extension",
    "VertApprAngle",
    "HorzApprAngle",
    "hitCSW",
    "hitCSW_resid"
)

data_filtered <- data %>%
  select(all_of(vars))

# Now to partition separate datasets for all the pitch types
data_partitioned <- list()

data_partitioned[["Fastball"]] <- data_filtered %>%
  filter(TaggedPitchType == "Fastball" | TaggedPitchType == "FourSeamFastBall" | TaggedPitchType == "OneSeamFastBall")
data_partitioned[["Curveball"]] <- data_filtered %>%
  filter(TaggedPitchType == "Curveball")
data_partitioned[["Slider"]] <- data_filtered %>%
  filter(TaggedPitchType == "Slider")
data_partitioned[["ChangeUp"]] <- data_filtered %>%
  filter(TaggedPitchType == "ChangeUp")
data_partitioned[["Sinker"]] <- data_filtered %>%
  filter(TaggedPitchType == "Sinker")

print(data_partitioned)
```

### Part 3a: Calculating Coefficients Using LM

Training Models

```{r}
# Model Variables used to predict Y, feel free to change
model_vars <- c(
    "RelSpeed", # Speed of ball at release
    "ZoneSpeed", # Speed of ball at the plate
    "EffectiveVelo", # Perceived pitch speed by the batter
    "VertBreak", # Full vertical break
    "InducedVertBreak", # Vertical break solely by the pitch(without the movement from gravity)
    "HorzBreak",
    "SpinRate",
    "SpinAxis", # spin axis from 0-360 degrees
    "RelHeight", # Vertical distance of the ball above home plate when the pitcher releases the balls
    "RelSide", # Distance from the y-axis from which the pitcher releases the ball
    "Extension", # Distance towards home plate from which the pitcher releases the ball relative to the pitching rubber
    "VertApprAngle", # Vertical Approach Angle - The vertical angle formed by the intersection of the y-axis and the ball’s path (in the z-direction) as it crosses the front of home plate. (Degrees)
    "HorzApprAngle" # Horizontal Approach Angle - The horizontal angle formed by the intersection of the y-axis and the ball’s path (in the x-direction) as it crosses the front of home plate (Degrees)
)

# get the LM equation formatted
equation <- paste("hitCSW_resid ~ ", paste(model_vars, collapse = " + "))
# print(equation)

# fit the model, predict stuff+ and standardize to 100
lms <- list()
stuff_plus <- list()
for(i in attr(data_partitioned, "names")) {
  cat(i, "Model: \n")
  lms[[i]] <- lm(formula(equation), data = data_partitioned[[i]])
  print(summary(lms[[i]]))
  cat("-------------------------------------------------\n")
  stuff_plus[[i]] <- data_partitioned[[i]] %>%
    mutate(
      raw_stuff = predict(lms[[i]], newdata = .),
      StuffPlus = 100 * raw_stuff / mean(raw_stuff, na.rm = TRUE)
    )
}
```


Ranking The Top 50 Pitches Given by Stuff Plus:

```{r}
for(i in attr(data_partitioned, "names")) {
  stuff_plus[[i]] %>%
    top_n(50, StuffPlus) %>%
    print(head(., n = 50))
}

```

```{r}
# Top Pitchers for each pitch

for(i in attr(data_partitioned, "names")) {
  stuff_plus[[i]] %>%
    group_by(Pitcher) %>%
    summarise(AvgStuffPlus = mean(StuffPlus, na.rm = TRUE)) %>%
    arrange(desc(AvgStuffPlus)) %>%
    mutate(Pitch = i) %>%
    print(head(., n = 50))
}
```


# Misc

## Stuff++ 

We could use a non-linear model to calculate Stuff+ but it would be a black-box model, meaning it gives us a score with no interpretable coefficients. The model might have better performance but low interpretability, hence we only know what the stuff is but we don't know what actually affects stuff.

ChatGPT Recommendation:

    Do I want to know the magnitude and direction of a variable’s impact?
    → Use lm, logistic, beta, GAM.

    Do I need to rank importance but not interpret exact relationships?
    → Try tree models + SHAP.

    Do I want interpretable nonlinear effects?
    → Use GAMs — they’re really underused for stuff like this
