---
title: "Baseball Pitch Analysis Pipeline"
execute:
  cache: true
format:
  pdf:
    geometry: left=0.3in, right=0.3in, top=0.3in, bottom=0.3in
    keep-tex: true
    include-in-header:
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
# Set random seed for reproducibility
set.seed(777)

# Load required libraries
library(tidyverse)    # Data manipulation and visualization
library(xgboost)      # Gradient boosting
library(caret)        # Machine learning tools
library(recipes)      # Feature engineering
library(mgcv)        # GAM models
library(ggplot2)     # Plotting
library(GGally)      # Correlation plots

# Parallel Computing Setup
library(parallel)
library(doParallel)
library(foreach)

# Detect number of CPU cores
num_cores <- detectCores()
cat("Detected", num_cores, "CPU cores\n")

# Set up parallel backend
# Use 75% of available cores to leave some for system processes
cores_to_use <- max(1, floor(num_cores * 0.75))
cat("Using", cores_to_use, "cores for parallel processing\n")

# Register parallel backend
cl <- makeCluster(cores_to_use)
registerDoParallel(cl)

# Set caret to use parallel processing
library(caret)
caret::registerDoParallel(cl)

cat("Parallel computing setup complete!\n")
cat("You can now run grid searches with parallel = TRUE\n")
```

# Data Loading and Initial Cleaning

## Load Raw Data
```{r load-data}
# Load data from different schools
ucla <- read_csv("./data/UCLA2023-2024.csv") %>%
  mutate(Date = as.Date(Date),
         UTCDate = as.Date(UTCDate),
         AwayTeamForeignID = as.character(AwayTeamForeignID))

penn <- read_csv("./data/PennState2024.csv")
purdue <- read_csv("./data/Purdue2024.csv")
michigan <- read_csv("./data/Michigan2024.csv")

# Combine all datasets
main <- bind_rows(ucla, penn, michigan, purdue)
```

## Select and Clean Variables
```{r clean-data}
# Define variables of interest
filtered_vars <- c(
  # Pitcher Information
  "Pitcher", "PitcherId", "PitcherThrows", "TaggedPitchType",
  
  # Velocity Metrics
  "RelSpeed", "ZoneSpeed", "EffectiveVelo", "ZoneTime", "SpeedDrop",
  
  # Movement Metrics
  "VertBreak", "InducedVertBreak", "HorzBreak",
  
  # Spin Metrics
  "SpinRate", "SpinAxis",
  
  # Release Metrics
  "RelHeight", "RelSide", "Extension",
  
  # Approach Angles
  "VertApprAngle", "HorzApprAngle", "VertRelAngle", "HorzRelAngle",
  
  # Location
  "PlateLocHeight", "PlateLocSide",
  
  # Game Situation
  "Balls", "Strikes", "Inning", "Outs",
  
  # Batter Information
  "Batter", "BatterTeam", "BatterSide",
  
  # Sequence Information
  "PAofInning", "PitchofPA", "PitchNo",
  
  # Quality Metrics
  "PitchReleaseConfidence", "PitchLocationConfidence", "PitchMovementConfidence"
)

# Check for missing values
missing_summary <- main %>%
  select(all_of(filtered_vars)) %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "na_count") %>%
  arrange(desc(na_count))

print("Missing value summary:")
print(missing_summary)

# Clean the dataset
main <- main %>%
  drop_na(all_of(filtered_vars)) %>%
  arrange(UTCDateTime, PitchNo)
```

# Feature Engineering

## Basic Transformations
```{r feature-engineering}
main <- main %>%
  # Create unique identifier
  mutate(DatasetID = 1:n()) %>%
  relocate(DatasetID, .before = PitchNo) %>%
  
  # Transform spin axis to circular components
  mutate(
    SpinAxis_rad = SpinAxis * pi / 180,
    SpinAxis_sin = sin(SpinAxis_rad),
    SpinAxis_cos = cos(SpinAxis_rad)
  ) %>%
  relocate(SpinAxis_rad, SpinAxis_sin, SpinAxis_cos, .after = SpinAxis) %>%
  
  # Create count factor and standardize release side
  mutate(
    Count = factor(paste0(Balls, "-", Strikes)),
    RelSide = abs(RelSide)  # Standardize release side for both RHP/LHP
  ) %>%
  relocate(Count, .after = Strikes)

# Save processed dataset
write_csv(main, "./data/stuff_data_full_raw.csv")
```

# Pitch-Specific Analysis

## Helper Functions
```{r helper-functions}
# Function to prepare approach angle data
prepare_approach_angles <- function(data) {
  # Vertical approach angle data
  vaa_df <- data %>%
    select(VertApprAngle, RelHeight, PlateLocHeight)
  vaa_recipe <- recipe(VertApprAngle ~ ., data = vaa_df) %>%
    step_normalize(all_numeric_predictors())
  vaa_prep <- prep(vaa_recipe)
  vaa_std <- bake(vaa_prep, new_data = NULL)
  
  # Horizontal approach angle data
  haa_df <- data %>%
    select(HorzApprAngle, RelSide, PlateLocSide)
  haa_recipe <- recipe(HorzApprAngle ~ ., data = haa_df) %>%
    step_normalize(all_numeric_predictors())
  haa_prep <- prep(haa_recipe)
  haa_std <- bake(haa_prep, new_data = NULL)
  
  list(
    vaa = vaa_std,
    haa = haa_std
  )
}

# Function to plot approach angle relationships
plot_approach_angles <- function(std_data, pitch_data, pitch_type) {
  std_data$vaa$PitcherSide <- factor(pitch_data$PitcherThrows)
  std_data$haa$PitcherSide <- factor(pitch_data$PitcherThrows)
  
  cat(paste("\nAnalyzing", pitch_type, "Approach Angles\n"))
  
  # Vertical approach angle plots
  print(ggpairs(std_data$vaa, 
                aes(color = PitcherSide),
                progress = FALSE,
                title = paste(pitch_type, "- Vertical Approach Angle")))
  
  # Horizontal approach angle plots
  print(ggpairs(std_data$haa, 
                aes(color = PitcherSide),
                progress = FALSE,
                title = paste(pitch_type, "- Horizontal Approach Angle")))
}

#' Create Tuning Grid for XGBoost Models
#' 
#' Creates a comprehensive tuning grid optimized for large datasets (~60k observations).
#' The grid is designed to balance computational efficiency with model performance.
#' 
#' @param grid_type Character string specifying the grid type:
#'   - "comprehensive": Full grid search (computationally expensive)
#'   - "balanced": Moderate grid search (recommended for 60k observations)
#'   - "quick": Minimal grid search (for testing)
#'   - "test": Ultra-minimal grid search (2-3 models for code testing)
#' @param n_obs Number of observations in the dataset (used to adjust grid size)
#' 
#' @return A data frame with tuning parameters for XGBoost
#' 
#' @examples
#' tune_grid <- create_xgb_tuning_grid("balanced", n_obs = 60000)
create_xgb_tuning_grid <- function(grid_type = "balanced", n_obs = 60000) {
  
  if (grid_type == "comprehensive") {
    # Full grid search - most thorough parameter exploration
    tune_grid <- expand.grid(
      nrounds = c(100, 200, 300, 500, 750, 1000),
      max_depth = c(2, 3, 4, 5, 6, 7, 8),
      eta = c(0.01, 0.03, 0.05, 0.1, 0.15, 0.2, 0.3),
      gamma = c(0, 0.1, 0.2, 0.5, 1.0),
      colsample_bytree = c(0.5, 0.6, 0.7, 0.8, 0.9, 1.0),
      min_child_weight = c(1, 2, 3, 5, 7, 10),
      subsample = c(0.5, 0.6, 0.7, 0.8, 0.9, 1.0)
    )
  } else if (grid_type == "balanced") {
    # Balanced grid - expanded for better performance
    tune_grid <- expand.grid(
      nrounds = c(100, 200, 300, 500),
      max_depth = c(3, 4, 5, 6, 7),
      eta = c(0.03, 0.05, 0.1, 0.15, 0.2),
      gamma = c(0, 0.1, 0.2, 0.5),
      colsample_bytree = c(0.6, 0.7, 0.8, 0.9, 1.0),
      min_child_weight = c(2, 3, 5, 7),
      subsample = c(0.6, 0.7, 0.8, 0.9, 1.0)
    )
  } else if (grid_type == "quick") {
    # Quick grid - for testing and development
    tune_grid <- expand.grid(
      nrounds = c(100, 200, 300),
      max_depth = c(3, 4, 5),
      eta = c(0.1, 0.2),
      gamma = c(0, 0.1),
      colsample_bytree = c(0.8, 1.0),
      min_child_weight = c(3, 5),
      subsample = c(0.8, 1.0)
    )
  } else if (grid_type == "test") {
    # Test grid - ultra-minimal for quick code testing (2-3 models)
    tune_grid <- expand.grid(
      nrounds = c(100, 200),
      max_depth = c(3, 4),
      eta = c(0.1),
      gamma = c(0),
      colsample_bytree = c(1.0),
      min_child_weight = c(5),
      subsample = c(0.8)
    )
  } else {
    stop("grid_type must be 'comprehensive', 'balanced', 'quick', or 'test'")
  }
  
  # Adjust grid size based on number of observations
  if (n_obs < 10000) {
    # For smaller datasets, reduce grid size
    tune_grid <- tune_grid[sample(nrow(tune_grid), min(30, nrow(tune_grid))), ]
  } else if (n_obs > 100000) {
    # For very large datasets, use comprehensive grid
    tune_grid <- expand.grid(
      nrounds = c(100, 200, 300, 500, 750),
      max_depth = c(3, 4, 5, 6, 7, 8),
      eta = c(0.01, 0.03, 0.05, 0.1, 0.15, 0.2),
      gamma = c(0, 0.1, 0.2, 0.5),
      colsample_bytree = c(0.6, 0.7, 0.8, 0.9, 1.0),
      min_child_weight = c(2, 3, 5, 7, 10),
      subsample = c(0.6, 0.7, 0.8, 0.9, 1.0)
    )
  }
  
  cat("Created tuning grid with", nrow(tune_grid), "combinations\n")
  return(tune_grid)
}

#' Create Train Control for Cross-Validation
#' 
#' Creates train control object optimized for the dataset size and computational resources.
#' 
#' @param cv_folds Number of cross-validation folds (default: 5 for 60k observations)
#' @param cv_repeats Number of CV repeats (default: 2 for stability)
#' @param verbose Logical. If TRUE, shows progress during training
#' @param parallel Logical. If TRUE, enables parallel processing
#' 
#' @return A trainControl object
#' 
#' @examples
#' train_control <- create_train_control(cv_folds = 5, verbose = FALSE)
create_train_control <- function(cv_folds = 5, cv_repeats = 2, verbose = FALSE, parallel = FALSE) {
  
  # Adjust CV folds based on dataset size
  if (cv_folds > 10) {
    warning("CV folds reduced to 10 for computational efficiency")
    cv_folds <- 10
  }
  
  train_control <- trainControl(
    method = "repeatedcv",
    number = cv_folds,
    repeats = cv_repeats,
    verboseIter = verbose,
    returnResamp = "all",
    savePredictions = TRUE,
    allowParallel = parallel
  )
  
  return(train_control)
}

#' Print Model Comparison Table
#' 
#' Creates and prints a comparison table of XGBoost model results.
#' 
#' @param model_results Results from caret train function
#' @param model_name Character string for the model name
#' @param top_n Number of top models to display
#' 
#' @return A formatted data frame with model comparisons
print_model_comparison <- function(model_results, model_name, top_n = 10) {
  
  # Extract results and add model name
  results_df <- model_results$results %>%
    mutate(Model = model_name) %>%
    arrange(RMSE) %>%
    select(Model, RMSE, Rsquared, MAE, nrounds, max_depth, eta, gamma, 
           colsample_bytree, min_child_weight, subsample)
  
  # Print summary
  cat(paste("\n=== Top", min(top_n, nrow(results_df)), "Models for", model_name, "===\n"))
  print(head(results_df, top_n))
  
  # Print best model details
  cat(paste("\nBest", model_name, "Model:\n"))
  best_model <- results_df[1, ]
  print(best_model)
  
  return(results_df)
}

#' Train Normalization Models for Fastballs
#' 
#' Trains GAM and XGBoost models for vertical approach angle normalization and XGBoost for horizontal approach angle.
#' Automatically selects the better performing model (GAM vs XGBoost) for vertical approach angle based on R-squared.
#' 
#' @param pitch_data A data frame containing pitch data with required columns:
#'   - VertApprAngle: Vertical approach angle
#'   - HorzApprAngle: Horizontal approach angle  
#'   - RelHeight: Release height
#'   - PlateLocHeight: Plate location height
#'   - RelSide: Release side (absolute value)
#'   - PlateLocSide: Plate location side
#' @param pitch_type Character string specifying the pitch type (e.g., "Fastball")
#' @param verbose Logical. If TRUE, prints detailed progress information and model summaries
#' @param grid_type Character string for tuning grid ("comprehensive", "balanced", "quick")
#' @param cv_folds Number of cross-validation folds
#' @param parallel Logical. If TRUE, enables parallel processing
#' 
#' @return A list containing:
#'   - data: data.frame - Original data with added normalized columns (HorzApprAngle_n, VertApprAngle_n)
#'   - vaa_model: The chosen model object (either GAM or XGBoost) for vertical approach angle
#'   - vaa_model_type: character - "GAM" or "XGBoost" indicating which model was selected
#'   - haa_xgb: train object - XGBoost model for horizontal approach angle
#'   - vaa_predictions: numeric vector - Predicted values for vertical approach angle
#'   - haa_predictions: numeric vector - Predicted values for horizontal approach angle
#'   - gam_r2: numeric - R-squared value for GAM model
#'   - xgb_r2: numeric - R-squared value for XGBoost model
#'   - tuning_results: data.frame - Cross-validation results for model selection
#'   - model_comparisons: list - Model comparison tables
#' 
#' @examples
#' fb_results <- train_pitch_normalization_fb(main_fb, "Fastball", verbose = FALSE, grid_type = "balanced")
#' main_fb <- fb_results$data
#' 
#' @importFrom mgcv gam
#' @importFrom caret train trainControl
#' @importFrom recipes recipe step_normalize prep bake
#' @importFrom dplyr select mutate relocate
train_pitch_normalization_fb <- function(pitch_data, pitch_type, verbose = FALSE, 
                                       grid_type = "balanced", cv_folds = 5, parallel = FALSE) {
  # Input validation
  required_cols <- c("VertApprAngle", "HorzApprAngle", "RelHeight", "PlateLocHeight", "RelSide", "PlateLocSide")
  missing_cols <- setdiff(required_cols, names(pitch_data))
  if (length(missing_cols) > 0) {
    stop("Missing required columns: ", paste(missing_cols, collapse = ", "))
  }
  
  if (nrow(pitch_data) == 0) {
    stop("pitch_data is empty")
  }
  
  if (verbose) cat(paste("\n=== Training normalization models for", pitch_type, "===\n"))
  if (verbose) cat("Dataset size:", nrow(pitch_data), "observations\n")
  
  # Prepare data for normalization
  vaa_df <- pitch_data %>%
    select(VertApprAngle, RelHeight, PlateLocHeight)
  
  haa_df <- pitch_data %>%
    select(HorzApprAngle, RelSide, PlateLocSide)
  
  # Normalize data for GAM
  rec <- recipe(VertApprAngle ~ ., data = vaa_df) %>%
    step_normalize(all_numeric_predictors())
  rec_prepped <- prep(rec)
  vaa_df_std <- bake(rec_prepped, new_data = NULL)
  
  # Train GAM model for vertical approach angle
  if (verbose) cat("Training GAM model for vertical approach angle...\n")
  vaa_gam <- gam(VertApprAngle ~ s(RelHeight) + PlateLocHeight, 
                 data = vaa_df_std %>% select(VertApprAngle, RelHeight, PlateLocHeight))
  
  # Create tuning grid and train control
  tune_grid <- create_xgb_tuning_grid(grid_type, n_obs = nrow(pitch_data))
  train_control <- create_train_control(cv_folds = cv_folds, verbose = FALSE, parallel = parallel)
  
  # Train XGBoost model for vertical approach angle
  if (verbose) cat("Training XGBoost model for vertical approach angle with", nrow(tune_grid), "parameter combinations...\n")
  vaa_xgb <- train(
    VertApprAngle ~ .,
    data = vaa_df,
    method = "xgbTree",
    trControl = train_control,
    tuneGrid = tune_grid,
    metric = "RMSE",
    verbosity = 0
  )
  
  # Compare R-squared values
  gam_r2 <- summary(vaa_gam)$r.sq
  xgb_r2 <- vaa_xgb$results$Rsquared[1]  # Get R-squared from CV results
  
  if (verbose) {
    cat("Model Comparison for Vertical Approach Angle:\n")
    cat("GAM R-squared:", round(gam_r2, 4), "\n")
    cat("XGBoost R-squared:", round(xgb_r2, 4), "\n")
    cat("Best XGBoost parameters:", paste(names(vaa_xgb$bestTune), vaa_xgb$bestTune, collapse = ", "), "\n")
    cat("Using", ifelse(xgb_r2 > gam_r2, "XGBoost", "GAM"), "model for vertical approach angle\n")
  }
  
  # Choose the better model for vertical approach angle
  if (xgb_r2 > gam_r2) {
    vaa_model <- vaa_xgb
    vaa_model_type <- "XGBoost"
    if (verbose) {
      cat("XGBoost Model Summary:\n")
      print(vaa_xgb)
    }
  } else {
    vaa_model <- vaa_gam
    vaa_model_type <- "GAM"
    if (verbose) {
      cat("GAM Model Summary:\n")
      print(summary(vaa_gam))
      plot(vaa_gam, pages = 1, main = paste(pitch_type, "- GAM Model"))
    }
  }
  
  # Train XGBoost model for horizontal approach angle
  if (verbose) cat("Training XGBoost model for horizontal approach angle...\n")
  haa_xgb <- train(
    HorzApprAngle ~ .,
    data = haa_df,
    method = "xgbTree",
    trControl = train_control,
    tuneGrid = tune_grid,
    metric = "RMSE",
    verbosity = 0
  )
  
  if (verbose) {
    cat("Horizontal XGBoost Model Summary:\n")
    print(haa_xgb)
    cat("Best horizontal XGBoost parameters:", paste(names(haa_xgb$bestTune), haa_xgb$bestTune, collapse = ", "), "\n")
  }
  
  # Generate model comparison tables
  vaa_comparison <- print_model_comparison(vaa_xgb, paste(pitch_type, "Vertical Approach Angle"), top_n = 10)
  haa_comparison <- print_model_comparison(haa_xgb, paste(pitch_type, "Horizontal Approach Angle"), top_n = 10)
  
  # Generate predictions
  if (verbose) cat("Generating predictions and creating normalized variables...\n")
  
  # Generate vertical predictions based on chosen model
  vert_hat <- predict(vaa_model, newdata = vaa_df %>% select(RelHeight, PlateLocHeight))
  horz_hat <- predict(haa_xgb, newdata = haa_df %>% select(RelSide, PlateLocSide))
  
  # Add normalized approach angles to dataset
  pitch_data_normalized <- pitch_data %>%
    mutate(
      HorzApprAngle_n = as.numeric(HorzApprAngle - horz_hat),
      VertApprAngle_n = as.numeric(VertApprAngle - vert_hat)
    ) %>%
    relocate(HorzApprAngle_n, VertApprAngle_n, .before = VertApprAngle)
  
  # Print summary statistics
  if (verbose) {
    cat("\nNormalized Value Summary:\n")
    cat("Horizontal Approach Angle Normalized:\n")
    print(summary(pitch_data_normalized$HorzApprAngle_n))
    cat("Vertical Approach Angle Normalized:\n")
    print(summary(pitch_data_normalized$VertApprAngle_n))
  }
  
  # Return list with normalized data and models
  return(list(
    data = pitch_data_normalized,
    vaa_model = vaa_model,
    vaa_model_type = vaa_model_type,
    haa_xgb = haa_xgb,
    vaa_predictions = vert_hat,
    haa_predictions = horz_hat,
    gam_r2 = gam_r2,
    xgb_r2 = xgb_r2,
    tuning_results = list(
      vaa_tuning = vaa_xgb$results,
      haa_tuning = haa_xgb$results
    ),
    model_comparisons = list(
      vaa_comparison = vaa_comparison,
      haa_comparison = haa_comparison
    )
  ))
}

#' Train Normalization Models for Breaking Balls (Sliders, Curveballs, Changeups)
#' 
#' Trains separate models for left-handed and right-handed pitchers for horizontal approach angle,
#' and automatically selects the better performing model (GAM vs XGBoost) for vertical approach angle.
#' This function handles the different trends between lefties and righties for breaking balls.
#' 
#' @param pitch_data A data frame containing pitch data with required columns:
#'   - VertApprAngle: Vertical approach angle
#'   - HorzApprAngle: Horizontal approach angle  
#'   - RelHeight: Release height
#'   - PlateLocHeight: Plate location height
#'   - RelSide: Release side (absolute value)
#'   - PlateLocSide: Plate location side
#'   - PitcherThrows: Pitcher handedness ("Left" or "Right")
#' @param pitch_type Character string specifying the pitch type (e.g., "Slider", "Curveball", "Changeup")
#' @param verbose Logical. If TRUE, prints detailed progress information and model summaries
#' @param grid_type Character string for tuning grid ("comprehensive", "balanced", "quick")
#' @param cv_folds Number of cross-validation folds
#' @param parallel Logical. If TRUE, enables parallel processing
#' 
#' @return A list containing:
#'   - data: data.frame - Original data with added normalized columns (HorzApprAngle_n, VertApprAngle_n)
#'   - vaa_model: The chosen model object (either GAM or XGBoost) for vertical approach angle
#'   - vaa_model_type: character - "GAM" or "XGBoost" indicating which model was selected
#'   - xgb_model_lefty: train object or NULL - XGBoost model for left-handed pitchers (NULL if no lefties)
#'   - xgb_model_righty: train object or NULL - XGBoost model for right-handed pitchers (NULL if no righties)
#'   - vaa_predictions: numeric vector - Predicted values for vertical approach angle
#'   - haa_predictions: numeric vector - Predicted values for horizontal approach angle
#'   - gam_r2: numeric - R-squared value for GAM model
#'   - xgb_r2: numeric - R-squared value for XGBoost model
#'   - tuning_results: list - Cross-validation results for all models
#'   - model_comparisons: list - Model comparison tables for all models
#' 
#' @examples
#' sl_results <- train_pitch_normalization_breaking_balls(main_sl, "Slider", verbose = FALSE, grid_type = "balanced")
#' main_sl <- sl_results$data
#' 
#' @importFrom mgcv gam
#' @importFrom caret train trainControl
#' @importFrom recipes recipe step_normalize prep bake
#' @importFrom dplyr select mutate relocate filter
train_pitch_normalization_breaking_balls <- function(pitch_data, pitch_type, verbose = FALSE, 
                                                   grid_type = "balanced", cv_folds = 5, parallel = FALSE) {
  # Input validation
  required_cols <- c("VertApprAngle", "HorzApprAngle", "RelHeight", "PlateLocHeight", 
                     "RelSide", "PlateLocSide", "PitcherThrows")
  missing_cols <- setdiff(required_cols, names(pitch_data))
  if (length(missing_cols) > 0) {
    stop("Missing required columns: ", paste(missing_cols, collapse = ", "))
  }
  
  if (nrow(pitch_data) == 0) {
    stop("pitch_data is empty")
  }
  
  # Check for valid pitcher handedness values
  valid_throws <- unique(pitch_data$PitcherThrows)
  if (!all(valid_throws %in% c("Left", "Right"))) {
    warning("Unexpected values in PitcherThrows: ", paste(valid_throws, collapse = ", "))
  }
  
  if (verbose) cat(paste("\n=== Training normalization models for", pitch_type, "===\n"))
  if (verbose) cat("Dataset size:", nrow(pitch_data), "observations\n")
  
  # Prepare data for normalization
  vaa_df <- pitch_data %>%
    select(VertApprAngle, RelHeight, PlateLocHeight)
  
  # Separate data for lefties and righties
  haa_lefty <- pitch_data %>%
    filter(PitcherThrows == "Left") %>%
    select(HorzApprAngle, RelSide, PlateLocSide)
  
  haa_righty <- pitch_data %>%
    filter(PitcherThrows == "Right") %>%
    select(HorzApprAngle, RelSide, PlateLocSide)
  
  if (verbose) {
    cat("Data split by pitcher handedness:\n")
    cat("Left-handed pitchers:", nrow(haa_lefty), "observations\n")
    cat("Right-handed pitchers:", nrow(haa_righty), "observations\n")
  }
  
  # Check if we have enough data for both groups
  if (nrow(haa_lefty) == 0 && nrow(haa_righty) == 0) {
    stop("No valid pitcher handedness data found")
  }
  
  # Normalize data for GAM
  rec <- recipe(VertApprAngle ~ ., data = vaa_df) %>%
    step_normalize(all_numeric_predictors())
  rec_prepped <- prep(rec)
  vaa_df_std <- bake(rec_prepped, new_data = NULL)
  
  # Train GAM model for vertical approach angle
  if (verbose) cat("Training GAM model for vertical approach angle...\n")
  vaa_gam <- gam(VertApprAngle ~ s(RelHeight) + PlateLocHeight, 
                 data = vaa_df_std %>% select(VertApprAngle, RelHeight, PlateLocHeight))
  
  # Create tuning grid and train control
  tune_grid <- create_xgb_tuning_grid(grid_type, n_obs = nrow(pitch_data))
  train_control <- create_train_control(cv_folds = cv_folds, verbose = FALSE, parallel = parallel)
  
  # Train XGBoost model for vertical approach angle
  if (verbose) cat("Training XGBoost model for vertical approach angle with", nrow(tune_grid), "parameter combinations...\n")
  vaa_xgb <- train(
    VertApprAngle ~ .,
    data = vaa_df,
    method = "xgbTree",
    trControl = train_control,
    tuneGrid = tune_grid,
    metric = "RMSE",
    verbosity = 0
  )
  
  # Compare R-squared values
  gam_r2 <- summary(vaa_gam)$r.sq
  xgb_r2 <- vaa_xgb$results$Rsquared[1]  # Get R-squared from CV results
  
  if (verbose) {
    cat("Model Comparison for Vertical Approach Angle:\n")
    cat("GAM R-squared:", round(gam_r2, 4), "\n")
    cat("XGBoost R-squared:", round(xgb_r2, 4), "\n")
    cat("Best XGBoost parameters:", paste(names(vaa_xgb$bestTune), vaa_xgb$bestTune, collapse = ", "), "\n")
    cat("Using", ifelse(xgb_r2 > gam_r2, "XGBoost", "GAM"), "model for vertical approach angle\n")
  }
  
  # Choose the better model for vertical approach angle
  if (xgb_r2 > gam_r2) {
    vaa_model <- vaa_xgb
    vaa_model_type <- "XGBoost"
    if (verbose) {
      cat("XGBoost Model Summary:\n")
      print(vaa_xgb)
    }
  } else {
    vaa_model <- vaa_gam
    vaa_model_type <- "GAM"
    if (verbose) {
      cat("GAM Model Summary:\n")
      print(summary(vaa_gam))
      plot(vaa_gam, pages = 1, main = paste(pitch_type, "- GAM Model"))
    }
  }
  
  # Train XGBoost models for horizontal approach angle (separate for lefties and righties)
  if (verbose) cat("Training XGBoost models for horizontal approach angle...\n")
  
  # Train model for lefties
  xgb_model_lefty <- NULL
  if (nrow(haa_lefty) > 0) {
    if (verbose) cat("Training XGBoost model for left-handed pitchers...\n")
    xgb_model_lefty <- train(
      HorzApprAngle ~ .,
      data = haa_lefty,
      method = "xgbTree",
      trControl = train_control,
      tuneGrid = tune_grid,
      metric = "RMSE",
      verbosity = 0
    )
    if (verbose) {
      cat("Left-handed XGBoost Model Summary:\n")
      print(xgb_model_lefty)
      cat("Best lefty XGBoost parameters:", paste(names(xgb_model_lefty$bestTune), xgb_model_lefty$bestTune, collapse = ", "), "\n")
    }
  }
  
  # Train model for righties
  xgb_model_righty <- NULL
  if (nrow(haa_righty) > 0) {
    if (verbose) cat("Training XGBoost model for right-handed pitchers...\n")
    xgb_model_righty <- train(
      HorzApprAngle ~ .,
      data = haa_righty,
      method = "xgbTree",
      trControl = train_control,
      tuneGrid = tune_grid,
      metric = "RMSE",
      verbosity = 0
    )
    if (verbose) {
      cat("Right-handed XGBoost Model Summary:\n")
      print(xgb_model_righty)
      cat("Best righty XGBoost parameters:", paste(names(xgb_model_righty$bestTune), xgb_model_righty$bestTune, collapse = ", "), "\n")
    }
  }
  
  # Generate model comparison tables
  vaa_comparison <- print_model_comparison(vaa_xgb, paste(pitch_type, "Vertical Approach Angle"), top_n = 10)
  
  model_comparisons <- list(vaa_comparison = vaa_comparison)
  
  if (!is.null(xgb_model_lefty)) {
    lefty_comparison <- print_model_comparison(xgb_model_lefty, paste(pitch_type, "Left-handed Horizontal"), top_n = 10)
    model_comparisons$lefty_comparison <- lefty_comparison
  }
  
  if (!is.null(xgb_model_righty)) {
    righty_comparison <- print_model_comparison(xgb_model_righty, paste(pitch_type, "Right-handed Horizontal"), top_n = 10)
    model_comparisons$righty_comparison <- righty_comparison
  }
  
  # Generate predictions
  if (verbose) cat("Generating predictions and creating normalized variables...\n")
  
  # Generate vertical predictions based on chosen model
  vert_hat <- predict(vaa_model, newdata = vaa_df %>% select(RelHeight, PlateLocHeight))
  
  # Generate horizontal predictions based on pitcher handedness
  horz_hat <- rep(NA, nrow(pitch_data))
  
  # Predict for lefties
  if (!is.null(xgb_model_lefty) && nrow(haa_lefty) > 0) {
    lefty_indices <- which(pitch_data$PitcherThrows == "Left")
    if (length(lefty_indices) > 0) {
      lefty_predictions <- predict(xgb_model_lefty, 
                                 newdata = pitch_data[lefty_indices, ] %>% 
                                   select(RelSide, PlateLocSide))
      horz_hat[lefty_indices] <- lefty_predictions
    }
  }
  
  # Predict for righties
  if (!is.null(xgb_model_righty) && nrow(haa_righty) > 0) {
    righty_indices <- which(pitch_data$PitcherThrows == "Right")
    if (length(righty_indices) > 0) {
      righty_predictions <- predict(xgb_model_righty, 
                                  newdata = pitch_data[righty_indices, ] %>% 
                                    select(RelSide, PlateLocSide))
      horz_hat[righty_indices] <- righty_predictions
    }
  }
  
  # Check if we have any NA predictions
  na_count <- sum(is.na(horz_hat))
  if (na_count > 0 && verbose) {
    warning(na_count, " horizontal predictions are NA - check pitcher handedness data")
  }
  
  # Add normalized approach angles to dataset
  pitch_data_normalized <- pitch_data %>%
    mutate(
      HorzApprAngle_n = as.numeric(HorzApprAngle - horz_hat),
      VertApprAngle_n = as.numeric(VertApprAngle - vert_hat)
    ) %>%
    relocate(HorzApprAngle_n, VertApprAngle_n, .before = VertApprAngle)
  
  # Print summary statistics
  if (verbose) {
    cat("\nNormalized Value Summary:\n")
    cat("Horizontal Approach Angle Normalized:\n")
    print(summary(pitch_data_normalized$HorzApprAngle_n))
    cat("Vertical Approach Angle Normalized:\n")
    print(summary(pitch_data_normalized$VertApprAngle_n))
  }
  
  # Return list with normalized data and models
  return(list(
    data = pitch_data_normalized,
    vaa_model = vaa_model,
    vaa_model_type = vaa_model_type,
    xgb_model_lefty = xgb_model_lefty,
    xgb_model_righty = xgb_model_righty,
    vaa_predictions = vert_hat,
    haa_predictions = horz_hat,
    gam_r2 = gam_r2,
    xgb_r2 = xgb_r2,
    tuning_results = list(
      vaa_tuning = vaa_xgb$results,
      lefty_tuning = if(!is.null(xgb_model_lefty)) xgb_model_lefty$results else NULL,
      righty_tuning = if(!is.null(xgb_model_righty)) xgb_model_righty$results else NULL
    ),
    model_comparisons = model_comparisons
  ))
}
```

## Analyze Each Pitch Type
```{r pitch-analysis}
# Summary of pitch types
pitch_summary <- main %>% 
  group_by(TaggedPitchType) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

print("Pitch type distribution:")
print(pitch_summary)

# Analyze Fastballs
main_fb <- main %>% 
  filter(TaggedPitchType %in% c("Fastball", "FourSeamFastBall"))
fb_angles <- prepare_approach_angles(main_fb)
plot_approach_angles(fb_angles, main_fb, "Fastball")

# Analyze Sliders
main_sl <- main %>% filter(TaggedPitchType == "Slider")
sl_angles <- prepare_approach_angles(main_sl)
plot_approach_angles(sl_angles, main_sl, "Slider")

# Analyze Curveballs
main_cv <- main %>% filter(TaggedPitchType == "Curveball")
cv_angles <- prepare_approach_angles(main_cv)
plot_approach_angles(cv_angles, main_cv, "Curveball")

# Analyze Changeups
main_chg <- main %>% filter(TaggedPitchType == "ChangeUp")
chg_angles <- prepare_approach_angles(main_chg)
plot_approach_angles(chg_angles, main_chg, "Changeup")
```

## Test Mode (Quick Code Validation)
```{r test-mode, eval=FALSE}
# Uncomment this section to test the code quickly
# This runs only 2-3 models per pitch type for quick validation

Test Fastball Normalization (2-3 models, ~5-10 minutes)
fb_test <- train_pitch_normalization_fb(
  main_fb,
  "Fastball",
  verbose = FALSE,
  grid_type = "test",
  cv_folds = 3,
  parallel = TRUE
)

# Test Slider Normalization (2-3 models, ~5-10 minutes)
sl_test <- train_pitch_normalization_breaking_balls(
  main_sl,
  "Slider",
  verbose = FALSE,
  grid_type = "test",
  cv_folds = 3,
  parallel = TRUE
)

# Test Curveball Normalization (2-3 models, ~5-10 minutes)
cv_test <- train_pitch_normalization_breaking_balls(
  main_cv,
  "Curveball",
  verbose = FALSE,
  grid_type = "test",
  cv_folds = 3,
  parallel = TRUE
)

# Test Changeup Normalization (2-3 models, ~5-10 minutes)
chg_test <- train_pitch_normalization_breaking_balls(
  main_chg,
  "Changeup",
  verbose = FALSE,
  grid_type = "test",
  cv_folds = 3,
  parallel = TRUE
)

cat("Test mode completed! If everything works, you can run the full pipeline.\n")
```

## Fastball Normalization
```{r fastball-normalization}
# Train normalization models for fastballs (same trend for lefties and righties)
# Using balanced grid search optimized for 60k observations with parallel processing
fb_results <- train_pitch_normalization_fb(
  main_fb, 
  "Fastball", 
  verbose = FALSE, 
  grid_type = "balanced", 
  cv_folds = 5, 
  parallel = TRUE
)
main_fb <- fb_results$data
```

## Slider Normalization
```{r slider-normalization}
# Train normalization models for sliders (separate models for lefties and righties)
# Using balanced grid search optimized for 60k observations with parallel processing
sl_results <- train_pitch_normalization_breaking_balls(
  main_sl, 
  "Slider", 
  verbose = FALSE, 
  grid_type = "balanced", 
  cv_folds = 5, 
  parallel = TRUE
)
main_sl <- sl_results$data
```

## Curveball Normalization
```{r curveball-normalization}
# Train normalization models for curveballs (separate models for lefties and righties)
# Using balanced grid search optimized for 60k observations with parallel processing
cv_results <- train_pitch_normalization_breaking_balls(
  main_cv, 
  "Curveball", 
  verbose = FALSE, 
  grid_type = "balanced", 
  cv_folds = 5, 
  parallel = TRUE
)
main_cv <- cv_results$data
```

## Changeup Normalization
```{r changeup-normalization}
# Train normalization models for changeups (separate models for lefties and righties)
# Using balanced grid search optimized for 60k observations with parallel processing
chg_results <- train_pitch_normalization_breaking_balls(
  main_chg, 
  "Changeup", 
  verbose = FALSE, 
  grid_type = "balanced", 
  cv_folds = 5, 
  parallel = TRUE
)
main_chg <- chg_results$data
```

# Save All Processed Datasets
```{r save-datasets}
# Save all pitch-specific datasets with normalizations
write_csv(main_fb, "./data/stuff_data_fastball_normalized.csv")
write_csv(main_sl, "./data/stuff_data_slider_normalized.csv")
write_csv(main_cv, "./data/stuff_data_curveball_normalized.csv")
write_csv(main_chg, "./data/stuff_data_changeup_normalized.csv")

cat("All datasets have been saved with normalizations applied.\n")
```

# Cleanup Parallel Computing
```{r cleanup, include=FALSE}
# Stop parallel cluster
if (exists("cl")) {
  stopCluster(cl)
  cat("Parallel cluster stopped.\n")
}

# Unregister parallel backend
unregisterDoParallel()
cat("Parallel computing cleanup complete.\n")
```
